---
title: "An Introduction to the printr Package"
author: "Yihui Xie"
date: "`r Sys.Date()`"
slug: printr
show_toc: yes
githubEditURL: https://github.com/yihui/printr/edit/master/vignettes/printr.Rmd
output:
  knitr:::html_vignette:
    number_sections: yes
    toc: yes
vignette: >
  %\VignetteIndexEntry{An Introduction to the printr Package}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
options(width = 80)
```

The **printr** (read "printer" or "print R") package is a companion package to
[**knitr**](https://yihui.name/knitr/). Its main purpose is to extend the S3
generic function `knit_print()` in **knitr**, which is the default value of the
chunk option `render`, as explained in the vignette
[knit_print.html](https://cran.rstudio.org/web/packages/knitr/vignettes/knit_print.html).

# Overview

To enable the printing methods defined in this package, just `library(printr)` or `loadNamespace('printr')`
in a code chunk (in the beginning) of your **knitr** document. Then some objects
will be printed differently with what you would have seen in a normal R console.
For example:

- matrices, data frames, and contingency tables are printed as tables (LaTeX,
  HTML, or Markdown, depending on your output format)
- the help page (from `?foo` or `help(foo)`) can be rendered as HTML, LaTeX, or
  plain text, and you can also specify which section(s) of the help page to
  include in the output
- the results from `browseVignettes()`, `help.search()`, `data()`, and
  `vignette()` are rendered as tables
- the package information from `library(help = 'foo')` is rendered as plain text

To disable the printing methods in this package, you can call
`detach('package:printr', unload = TRUE)` if you _attached_ the package via
`library(printr)` before, or `unloadNamespace('printr')` if you _loaded_ it via
`loadNamespace('printr')`.

This package aims to be portable in the sense that it should work in most
document formats, including `*.Rnw` (R + LaTeX), `*.Rmd` (R Markdown), and
`*.Rhtml` (R + HTML) files, etc.

You can find the package source as well as installation instructions on [Github](https://github.com/yihui/printr),
and you are welcome to contribute code via pull requests, or file feature
requests and bug reports via [Github
issues](https://github.com/yihui/printr/issues).

# Examples

First we take a look at a quick example of printing some R objects in the R
console:

```{r comment='', prompt=TRUE}
# R uses plain text representation for data frames/matrices/...
head(mtcars)
head(iris)
```

Then we attach the **printr** package in this R session, and see how things
change later:

```{r}
library(printr)
```

## Matrices/data frames/tables

Matrices and data frames are printed as tables using the `kable()` function in
**knitr**:

```{r warning=FALSE}
options(digits = 4)
set.seed(123)
x = matrix(rnorm(40), 5)
x
# with colunm names
dimnames(x) = list(NULL, head(LETTERS, ncol(x)))
x
# further customization via kable(), e.g. digits and captions
knitr::kable(x, digits = 2, caption = 'A table produced by printr.')
head(mtcars)
head(iris, 10)
```

For contingency tables, 1-d tables are printed as a 1-row matrix, 2-d tables are
printed an $n \times m$ matrix, and tables of higher dimensions are printed as
data frames with frequencies.

```{r}
x1 = sample(letters[1:3], 1000, TRUE)
x2 = sample(letters[1:3], 1000, TRUE)
x3 = sample(letters[1:3], 1000, TRUE)
table(x1)
table(x1, x2)
table(x1, x2, x3)
```

## Search results from `help.search()`

Here are some examples demonstrating the results of `help.search()`, or you can
also use `??` to search for a string.

```{r tidy=FALSE}
??sunflower
help.search('contourplot')
help.search('foo', package = 'base')
help.search('foooooooo', package = 'utils')
```

In a normal R session, the results will be displayed as an HTML page by default,
but normally these functions are meant to be called in an interactive R session,
and **knitr** documents are often compiled in non-interactive R sessions, so we
changed the printing behavior of these results, and readers will get the basic
idea of these functions when reading the **knitr** output. If they want to run
these functions by themselves, they can do it in an interactive R session.

## Help pages

When you want to read the help page of a certain R object, you normally use `?`
or `help()`, which will launch a separate help page from the R session, and
require human interaction. Again, we may not desire human interactions in
**knitr** documents, so the help pages are printed as static documents here.

```{r tidy=FALSE, comment=''}
?coef
```

When help pages are really long, we can use the chunk option
`printr.help.sections` to select
a few sections to display, e.g. we only show the sections `description` and
`usage` of the `paste()` function:

```{r tidy=FALSE, printr.help.sections=c('description', 'usage'), comment=''}
?paste
```

## Vignette/dataset lists

We can print the lists of vignettes and datasets in packages using `vignette()`
and `data()`, respectively.

```{r error=TRUE, purl=FALSE}
vignette(package = 'rpart')
vignette(package = c('rpart', 'knitr'))
data(package = 'lattice')
data(package = c('rpart', 'lattice'))
data(package = 'knitr')  # no datasets here
browseVignettes(package = 'knitr')
```

## Package info

A description of a package can be printed via `library(help = 'foo')`:

```{r comment='', R.options=list(width=100)}
library(help = 'printr')
```
